---
title: "Reduce dimensionality"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pkg}
library(tidyverse)
library(xlsx)
```


```{r}
d <- read.xlsx("rawdata/Tabla_variables_modelización_lepidópteros_ACTUALIZADO_2019.xls", 1) %>% 
  rename(code = Abreviatura)
```

#### Average climate variables

```{r}
# Precipitacion Acumulada 
precWI <- d %>% 
  dplyr::select(code, contains("Precipitación.Invierno")) %>% 
  dplyr::select(code, contains("Acumulado")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, precWI = m)

precAU <- d %>% 
  dplyr::select(code, contains("Precipitación.Otoño")) %>% 
  dplyr::select(code, contains("Acumulado")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, precAU = m)

precSP <- d %>% 
  dplyr::select(code, contains("Precipitación.Primavera")) %>% 
  dplyr::select(code, contains("Acumulado")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, precSP = m)

precSU <- d %>% 
  dplyr::select(code, contains("Precipitación.Verano")) %>% 
  dplyr::select(code, contains("Acumulado")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, precSU = m)

precYE <- d %>% 
  dplyr::select(code, contains("Acumulado.ANUAL")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, precYE = m)
  

prec <- precYE %>% 
  inner_join(precAU) %>% 
  inner_join(precWI) %>% 
  inner_join(precSP) %>% 
  inner_join(precSU) 

```

```{r}
# Temperatura media 

tempWI <- d %>% 
  dplyr::select(code, contains("Temperatura.Invierno")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, tempWI = m)

tempAU <- d %>% 
  dplyr::select(code, contains("Temperatura.Otoño")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, tempAU = m)

tempSP <- d %>% 
  dplyr::select(code, contains("Temperatura.Primavera")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, tempSP = m)

tempSU <- d %>% 
  dplyr::select(code, contains("Temperatura.Verano")) %>% 
  rowwise() %>% 
  mutate(m = mean(c_across(where(is.numeric)))) %>% 
  dplyr::select(code, tempSU = m)

temp <- tempAU %>% 
  inner_join(tempWI) %>% 
  inner_join(tempSP) %>% 
  inner_join(tempSU) 

```

```{r}
df  <- d %>% 
  dplyr::select(code, elev = Elevación, TP_ES_OE:LS_NDVI) %>% 
  inner_join(prec) %>% 
  inner_join(temp)
```


### Multicollinearity
- Correlation matrix. When computing a matrix of Pearson’s bivariate correlations among all independent variables, the magnitude of the correlation coefficients should be less than .80

- Variance Inflation Factor (VIF) – The VIFs of the linear regression indicate the degree that the variances in the regression estimates are increased due to multicollinearity. VIF values higher than 10 indicate that multicollinearity is a problem.

See: 
- https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0202403 


In practice, the rule‐of‐thumb method in dealing with collinearity is to minimize its potential effect by selecting variables whose correlation coefficients are below a certain threshold (e.g., |r| <0.7 in Dormann et al. (2013) or <0.4 in Suzuki, Olson, & Reilly (2008)).
https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.5555 



Initially, a hierarchical correlation dendrogram was used to find groups of highly correlated variables (midrmid > 0.95) with the varclus function from 'Hmisc' package (Harrell Jr 2019). For each of these clusters, the variable with the lowest correlation (i.e. less redundancy) with all variables present in the data set was selected.


Methods for dealing with collinearity (from Dormann et al 2012): 
As a general rule of thumb, a good strategy is to select variables that:
- are ecologically relevant
- are feasible to collect data on
- are closer to the mechanism (in the sequence resource-direct-indirect-proxy variables: Harrell 2001, Austin 2002).

### Correlation
```{r}
# http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software
library(Hmisc)
res2 <- rcorr(as.matrix(df[,-1]))



# Assess p value 
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

cop <- flattenCorrMatrix(res2$r, res2$P)
```

```{r}
### Visuazize 
res <- cor(df[,-1])
s <- symnum(res, abbr.colnames = FALSE)


library(corrplot)
cor.mtest <- function(mat, ...) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

p.mat <- cor.mtest(df[,-1])


corrplot(res, type = "upper", order = "hclust", tl.col = "black", tl.srt = 90, tl.cex = 0.4,
         p.mat = p.mat, sig.level = 0.1, 
         insig = "blank", 
         diag = FALSE)
```

```{r}
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = res, col = col, symm = TRUE)
```


# PCA 

```{r}
library("nFactors")
library("factoextra")
pca <- prcomp(df[,-1])

```

```{r}
fviz_eig(pca)
plot(nScree((pca$sdev)^2))

pca_biplot <- fviz_pca_var(pca,
                          col.var = "contrib", # Color by contributions to the PC
                          repel = TRUE     # Avoid text overlapping
) + scale_color_gradient(low="blue", high="green") + theme_bw()








# Get Eigenvalues (sd^2)
ev <- (pca$sdev)^2

# Another way 
ev1 <- get_eigenvalue(pca)

# Component importance 
compImp <- summary(p_env)
compImp <- t(as.data.frame(compImp$importance))

# Loadings 
pcaLoadings <- as.data.frame(round(pca$rotation[,c(1:3)], 3))
pcaLoadings$variables <- rownames(p_envLoadings) 
rownames(p_envLoadings) <- NULL

# Correlation Variables-axis
p_env_var <- get_pca_var(p_env)
p_envCor <- as.data.frame(round(p_env_var$cor[, c(1:3)], 3))
p_envCor$variables <- rownames(p_envCor)
rownames(p_envCor) <- NULL







```


